### 2024-01-10 inst/scripts/papertest.Rmd
paper experimental setup

## A simple vignette to explore simple neutral transmission between realtively stable, and close to equilibrium population

```{r}
devtools::load_all(".")
```

We have z neutral traits, specifc for each community.

```{r}
z=4
neutraltraitsParam=initNeutralTraitsPathways(z = z)
neutraltraitsParam$post[,"i"]=1
neutraltraitsParam$s=rep(.5,z)
traitsid=paste0("t",1:z)
```

To simplify we define $K=z$ communities, all communities of hunter gatherer , with 150 individual per community, 150 will also be the Fission Threshold (ie community starts at maximum pop size).
```{r initparam,out.width="40%"}
km=0
ki=z
G=2
initcomus=initialiseCommunities(traits=initAdaptiveTraits(ki=ki,km=km,n=20),coordinates=random2Dgrid(K=ki+km,Gx=G),G=G,sizes=150,plot=T)
text(initcomus$coordinates[,1],initcomus$coordinates[,2],labels=1:nrow(initcomus$coordinates))
cols=c("#E41A1C","#377EB8","#4DAF4A","#984EA3")
plot(initcomus$coordinates[,1],initcomus$coordinates[,2],bg=cols,pch=21,cex=20)
text(initcomus$coordinates[,1],initcomus$coordinates[,2],labels=1:nrow(initcomus$coordinates))
```

Age of people is initially random
```{r randompop}
communities=unlist(lapply(seq_along(initcomus$size),function(i)rep(i,initcomus$size[i])))
population=cbind(newpop(n=sum(initcomus$size),age="random",community = communities),initNeutralTraits(sum(initcomus$size),z))
```

Each community will has a one specific neutral traits

```{r assigntraits,results='asis'}
population[,traitsid]=0
for(i in 1:z){population[population[,"community"]==i,traitsid[i]]=1}
knitr::kable(aggregate(population[,traitsid],by=list(population[,"community"]),sum))
```

```{r runsimu,warning=FALSE,message=FALSE,results='hold',cache=TRUE}
tstep=10000
quickV=modelVector(K=ki, m=1, b=0.2, r=0.01, rho=0, d=0.01, maturity=18, endrepro=65, population=population, comus=initcomus, tstep=tstep, tp=neutraltraitsParam,age.threshold=20, out=c("finalpop","finalcomus","popsize"),logging=c("time"),ma=1,traitsid=paste0("t",1:z),F_Th = 150)
data(population)

@@

check that the population is stable:

```{r followpopsize}
plot(quickV$popsize,type="l",ylim=c(300,650),lwd=5,xlab="time",ylab="number of individual")
```

What is the distribution of traits at the end:

```{r finaldistrib,fig.show="hold"}
knitr::kable(aggregate(quickV$population[,traitsid],by=list(quickV$population[,"community"]),FUN=getRatio))

barplot(as.matrix(aggregate(quickV$population[,traitsid],by=list(quickV$population[,"community"]),FUN=getRatio)[,-1]),beside=T,col=cols,main="percentage of population with traits")
barplot(table(quickV$population[,"community"]),col=cols,main="community size")

```
What if it's horizontal transmission pre-marital only? Expecting less changes?
```{r runsimu-pre,warnings=FALSE,results=FALSE,results='hold',cache=TRUE}
neutraltraitsParam$post[,"i"]=0
neutraltraitsParam$pre[,"h"]=1
quickV=modelVector(K=ki, m=1, b=0.2, r=0.01, rho=0, d=0.01, maturity=18, endrepro=65, population=population, comus=initcomus, tstep=tstep, tp=neutraltraitsParam,age.threshold=20, out=c("finalpop","finalcomus","popsize"),logging=c(""),ma=1,traitsid=paste0("t",1:z),F_Th = 150)
plot(quickV$popsize,type="l",ylim=c(300,650),lwd=5,xlab="time",ylab="number of individual")
knitr::kable(aggregate(quickV$population[,traitsid],by=list(quickV$population[,"community"]),FUN=getRatio))

cols=c("#E41A1C","#377EB8","#4DAF4A","#984EA3")
barplot(as.matrix(aggregate(quickV$population[,traitsid],by=list(quickV$population[,"community"]),FUN=getRatio)[,-1]),beside=T,col=cols,main="percentage of population with traits")
barplot(table(quickV$population[,"community"]),col=cols,main="community size")

@@

```{r,eval=F}
tstep=500
repli=50
library(parallel)
neutraltraitsParam$s=c(.5,.5,.5,.5)
cl <- makeCluster(10,type="FORK")

rhodep=lapply(c(0,.5,1),function(rho){
           neutraltraitsParam$post[,]=0
           neutraltraitsParam$pre[,]=0
           neutraltraitsParam$post[,"i"]=1
           #allinlaw=replicate(repli,
           allinlaw=parSapply(cl,1:repli,function(r)
                              {
                                  set.seed(as.numeric(Sys.time())+r)
                                  quickV=modelVector(K=K, m=1, b=0.2, r=0.01, rho=rho, d=0.01, maturity=18, endrepro=65, population=population, comus=initcomus, tstep=tstep, tp=neutraltraitsParam,age.threshold=20, out=c("finalpop","finalcomus","popsize"),logging=c("time"),ma=1,traitsid=paste0("t",1:z),F_Th = 150,vidfile=NULL)
                                  pop=quickV$population
                                  #aggregate(pop[,traitsid],by=list(pop[,"community"]),sum)[-1,]
                                  sapply(1:z,function(i)apply(pop[pop[,"community"]==i,traitsid],2,sum))
                              },simplify=F)

           neutraltraitsParam$post[,"i"]=0
           neutraltraitsParam$post[,"h"]=1
           #allposthoriz=replicate(repli,{
           allposthoriz=parSapply(cl,1:repli,function(r){
                                  set.seed(as.numeric(Sys.time())+r)
                                      quickV=modelVector(K=K, m=1, b=0.2, r=0.01, rho=rho, d=0.01, maturity=18, endrepro=65, population=population, comus=initcomus, tstep=tstep, tp=neutraltraitsParam,age.threshold=20, out=c("finalpop","finalcomus","popsize"),logging=c("time"),ma=1,traitsid=paste0("t",1:z),F_Th = 150,vidfile=NULL)
                                      pop=quickV$population
                                      #aggregate(pop[,traitsid],by=list(pop[,"community"]),sum)
                                      sapply(1:z,function(i)apply(pop[pop[,"community"]==i,traitsid],2,sum))
                              },simplify=F)

           neutraltraitsParam$post[,"h"]=0
           neutraltraitsParam$pre[,"h"]=1
           #allprehoriz=replicate(repli,{
           allprehoriz=parSapply(cl,1:repli,function(r){
                                  set.seed(as.numeric(Sys.time())+r)
                                     quickV=modelVector(K=K, m=1, b=0.2, r=0.01, rho=rho, d=0.01, maturity=18, endrepro=65, population=population, comus=initcomus, tstep=tstep, tp=neutraltraitsParam,age.threshold=20, out=c("finalpop","finalcomus","popsize"),logging=c("time"),ma=1,traitsid=paste0("t",1:z),F_Th = 150,vidfile=NULL)
                                     pop=quickV$population
                                     #aggregate(pop[,traitsid],by=list(pop[,"community"]),sum)
                                     sapply(1:z,function(i)apply(pop[pop[,"community"]==i,traitsid],2,sum))
                              },simplify=F)

@@

stopCluster(cl)
saveRDS(file="precomp.RDS",rhodep)
```

>>>>

Attempt on realising the experiment on the paper:

```{r}
z=24
neutraltraitsParam=initNeutralTraitsPathways(z = z)

pw=1
for(sb in c(.5,1)){
    print(pw)
    neutraltraitsParam$s[pw]=sb
    for(pre in c("v","h","o")){
        neutraltraitsParam$pre[pw:(pw+3),pre]=1
        pw=pw+1
        print(pw)
        neutraltraitsParam$s[pw]=sb
        for(post in c("h","o","i")){
            neutraltraitsParam$post[pw,post]=1
            print(pw)
            neutraltraitsParam$s[pw]=sb
            pw=pw+1
        }
    }
}

print.pathways(neutraltraitsParam,verb=F)

@@

adaptive traits and communities:
```{r}
km=5
ki=z-km
traits=initAdaptiveTraits(ki=ki,km=km,n=3) # this generate a table for 3 adaptive traits, with ki communities of full hunter-gathere and km of farmers

G=5 #limits/numer of row & column of the world grid
coordinates=random2Dgrid(K=ki+km,Gx=G) #random coordinated on the grid

initcomus=initialiseCommunities(traits=traits,coordinates=coordinates,G=G,sizes=100,plot=T)

communities=unlist(lapply(seq_along(initcomus$size),function(i)rep(i,initcomus$size[i])))
population=cbind(newpop(n=sum(initcomus$size),age="random",community = communities),initNeutralTraits(sum(initcomus$size),z))
```

We run a first simulation to have a proper age and family structure

```{r runburnin,warning=FALSE,message=FALSE,results='hold',cache=TRUE}
tstep=500
burnin=modelVector(
   K=ki, m=1, b=0.2, r=0, rho=0, d=0.01, maturity=18, endrepro=65, 
   population=population, 
   comus=initcomus, 
   tstep=tstep, 
   tp=initNeutralTraitsPathways(z = z) , #we recreate a temporary neutral trait table with no neutral transmission
   age.threshold=20, 
   out=c("finalpop","finalcomus"),   #we only want the final population
   logging=c("time"),ma=1,traitsid=paste0("t",1:z),F_Th = 100
)
population=burnin$population
initcomus$size=burnin$finalcomus$size
```

```{r setuppopulation}
traitsid=paste0("t",1:z)
population[,traitsid]=0
for(i in 1:z){population[population[,"community"]==i,traitsid[i]]=1}
knitr::kable(aggregate(population[,traitsid],by=list(population[,"community"]),sum))
```

We can also reset ids to start at one and have more readable results:

```{r resetids}
population[population[,"cid"]!= -1 ,"cid"] = as.numeric(as.factor(as.character(population[population[,"cid"]!= -1 ,"cid"])))
population[,"fid"] = as.numeric(as.factor(as.character(population[ ,"fid"])))
ids = as.numeric(as.factor(as.character(population[ ,"id"])))
names(ids)=population[,"id"]
population[,"id"] = as.numeric(as.factor(as.character(population[ ,"id"])))
population[,"partner"]=as.numeric(ids[as.character(population[,"partner"])])
population[is.na(population[,"partner"]),"partner"]=-1
```

```{r runpropersimu}
tstep=1000
a=Sys.time()
initcomus$occupation[,]=1 #cheat and prohibit the empty space to be colonised
simuTest=modelVector(
    K=ki, m=1, b=0.2, r=0.01, rho=0, d=0.01, 
    maturity=18, endrepro=65, 
    population=population, 
    comus=initcomus, 
    tstep=tstep, 
    tp=neutraltraitsParam,
    age.threshold=20, 
    out=c("finalpop","finalcomus","traitsumary","popsize"), #this time we keep track of more things
    logging=c("time","done","visu"),
    beta=-1,
    ma=1,traitsid=paste0("t",1:z),F_Th = 150
)
print(Sys.time()-a)
```

The choice of parameter should ensure a stable population:

```{r followpopsize}
plot(simuTest$popsize,type="l",ylim=c(300,650),lwd=5,xlab="time",ylab="number of individual")
```

The parameter `out` define what's available as outcome, and multiple things are available. `finalpop` tells the model to return the population at the final time step. Using that we can look at the distribution of traits at the end:

```{r finaldistrib,fig.show="hold"}

knitr::kable(aggregate(simuTest$population[,traitsid],by=list(simuTest$population[,"community"]),FUN=getRatio))

barplot(as.matrix(aggregate(simuTest$population[,traitsid],by=list(simuTest$population[,"community"]),FUN=getRatio)[,-1]),beside=T,col=cols,main="percentage of population with traits")
barplot(table(simuTest$population[,"community"]),col=cols,main="community size")

simuTest$finalcomus

We can track the frequencie of each traits using "traitsumary"

```{r runsimu-pre,warnings=FALSE,results=FALSE,results='hold'}

@@

```{r,eval=F}
tstep=500
repli=50
neutraltraitsParam$s=c(.5,.5,.5,.5)
rhodep=lapply(c(0,1),function(rho){
           neutraltraitsParam$post[,]=0
           neutraltraitsParam$pre[,]=0
           neutraltraitsParam$post[,"i"]=1
           allinlaw=replicate(repli,
                              {
                                  quickV=modelVector(K=K, m=1, b=0.2, r=0.01, rho=rho, d=0.01, maturity=18, endrepro=65, population=population, comus=initcomus, tstep=tstep, tp=neutraltraitsParam,age.threshold=20, out=c("finalpop","finalcomus","popsize"),logging=c("time"),ma=1,traitsid=paste0("t",1:z),F_Th = 150,vidfile=NULL)
                                  pop=quickV$population
                                  sapply(1:z,function(i)apply(pop[pop[,"community"]==i,traitsid],2,sum))
                              },simplify=F)

           neutraltraitsParam$post[,"i"]=0
           neutraltraitsParam$post[,"h"]=1
           allposthoriz=replicate(repli,{
                                  set.seed(as.numeric(Sys.time())+r)
                                      quickV=modelVector(K=K, m=1, b=0.2, r=0.01, rho=rho, d=0.01, maturity=18, endrepro=65, population=population, comus=initcomus, tstep=tstep, tp=neutraltraitsParam,age.threshold=20, out=c("finalpop","finalcomus","popsize"),logging=c("time"),ma=1,traitsid=paste0("t",1:z),F_Th = 150,vidfile=NULL)
                                      pop=quickV$population
                                      sapply(1:z,function(i)apply(pop[pop[,"community"]==i,traitsid],2,sum))
                              },simplify=F)

           neutraltraitsParam$post[,"h"]=0
           neutraltraitsParam$pre[,"h"]=1
           allprehoriz=replicate(repli,{
                                  set.seed(as.numeric(Sys.time())+r)
                                     quickV=modelVector(K=K, m=1, b=0.2, r=0.01, rho=rho, d=0.01, maturity=18, endrepro=65, population=population, comus=initcomus, tstep=tstep, tp=neutraltraitsParam,age.threshold=20, out=c("finalpop","finalcomus","popsize"),logging=c("time"),ma=1,traitsid=paste0("t",1:z),F_Th = 150,vidfile=NULL)
                                     pop=quickV$population
                                     sapply(1:z,function(i)apply(pop[pop[,"community"]==i,traitsid],2,sum))
                              },simplify=F)

@@
